A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-02-10 papers/contact/paper.tex
clean up contact paper, and avoid re-running mc calculations Tested everything on bingley.
various changes to papers Tested everything on bingley.

\author{David Roundy}
\affiliation{Department of Physics, Oregon State University, Corvallis, OR 97331}

@@

There has been considerable recent interest in using Statistical
Associating Fluid Theory (SAFT) to construct classical density
functionals to describe associating
fluids\cite{\saftlocaldft,\saftnonlocaldft}.  This approach has been
successful in qualitatively describing the dependence of surface
tension on temperature.  Unfortunately, most of these constructed
functionals\cite{\saftlocaldft} cannot be used to study arbitrary
inhomogeneous density distributions, because they use local
density approximations that are only valid for densities that vary
slowly over molecular length scales.

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  Each of these terms has some dependence on the
inhomogeneity in the density.  The ideal gas is purely local, and in
fact must be the only purely local term in the free energy in order
for the functional to satisfy the contact-value theorem.  The
hard-sphere contribution contribution to the free energy has been
thoroughly studied, and is well approximated by the White Bear version
of the Fundamental-Measure Theory (FMT)
functional\cite{roth2002whitebear}.  A functional for the chain
contribution is has recently been introduced\cite{bryk2006density,
  gross2009density}, but is not needed for small associating molecules
such as water.  The dispersion term is long-range and thus has
significant dependence on the density distribution, but because its
relatively weak position dependence, is amenable to mean-field
approximations.  A functional for the association term of the free
energy was introduced by Wu and
Yu\cite{yu2002fmt-dft-inhomogeneous-associating}, and was used in
several subsequent functionals\cite{fu2005vapor-liquid-dft,
  bryk2006density}.  A key input (FIXME: term, factor?) to both the
association and chain free energies is the contact value of the
correlation function of a hard-sphere fluid.

The contact value of the correlation function---which is the same as
the contact value of the cavity correlation function for hard
spheres---is the the value of the correlation function evaluated a
distance equal to the hard-sphere diameter.  In an inhomogeneous
system treated using fundamental measures, there is some ambiguity in
the meaning of the \emph{local} contact value of the correlation
function which appears in the equations for the homogeneous system.
Clearly the contact value of the correlation function will depend on
position, but how should that position be defined? It will depend on
how the correlation function is used.  There are two reasonable
choices for functionals based on fundamental measure theory, which
will depend on which densities are used...
\begin{align}
  F[n(\rr)] = \int n(\rr) f(\bar{n}(\rr), g_\sigma(\rr)) d\rr
\end{align}
\begin{align}
  F[n(\rr)] = \int f(n_2(\rr), g_\sigma(\rr)) d\rr
\end{align}

Yu and Wu introduce an approximation based on the fundamental measures
of Fundamental Measure Theory (see Section~\ref{sec:yuwu}).  Gross
also does\cite{gross2009density}.

In this paper, we will report on the accuracy of this approximation by
comparing with Monte Carlo simulations of the inhomogeneous
hard-sphere fluid.  In addition, we will introduce and test differing
approximations for the cavity correlation function.

\textcolor{red}{Cavity correlation function!}

The hard-sphere contact density---that is to say, the density of hard
spheres touching a given hard sphere---is important for the
association term in Statistical Associating Fluid Theory (SAFT).
For a homogeneous hard-sphere fluid, the contact density---which is to
say, $n g_{HS}(\sigma)$, where $g_{HS}(\sigma)$ is the contact value
of the correlation function---is well-known.  However, for
inhomogeneous fluids, the contact density has not been well studied.

{
  \color{blue}We need to figure out a few bits of terminology.

  \begin{enumerate}
  \item What should we call the contact density between two hard
    spheres, to distinguish it from the contact density at any other
    hard surface.  One option would be to focus on the contact value
    of the correlation function, which would make it clear what we're
    talking about.
  \item How should we distinguish between the various definitions.  In
    particular, ``at the sphere'' vs. ``at contact''.  It'd be helpful
    to have a short way of describing these.
  \end{enumerate}
}

@@

\section{Direct correlation function}

\textcolor{red}{FIXME WRITE THIS SECTION!} The direct correlation
function is readily accessible from density-functional theory, as the
second functional derivative of the free energy:
\textcolor{red}{(Something like...)}
\begin{align}
  -c^{(2)}(\xx_1, \xx_2) &= \frac{1}{k_BT}\frac{\delta^2 F_{ex}}{\delta n(\xx_1) \delta n(\xx_2)}
\end{align}
The direct correlation function is related to the ordinary correlation
function by the Ornstein-Zernike relation
\begin{align}
  h(\xx_1, \xx'_2) &= g(\xx_1, \xx_2) - 1 \\
  h(\xx_1, \xx_2) &= c(\xx_1, \xx_2)
    + \int n(\xx_3)c(\xx_1, \xx_3) h(\xx_2, \xx_3)d\xx''
\end{align}
From these two equation, we could in principle extract the correlation
function evaluated at contact from the DFT free energy functional.
The Ornstein-Zernike equation is usually solved by taking a Fourier
transform of all the relevant quantities.  In the inhomogeneous case,
however, this approach isn't workable, since the Fourier transforms
don't speed things up like you'd hope, and we don't have a periodic
system in general.

\textcolor{red}{It seems worth looking into this, to see what (if
  anything) we can do! I don't see a solution just now, but that
  doesn't mean there isn't one around somewhere.  In any case, just
  looking at the direct correlation function might give us some
  hints as to a different and interesting approximation for the
  contact density.}

@@

\section{Mean Contact density}\label{simple-contact}

One reasonably simple question to ask is what the \emph{mean} contact
density of an inhomogeneous system is.  This comes down to simply
reproducing the derivation for a homogeneous system using the
FMT functional.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  \ncontact &= \frac{1}{N k_BT 4\pi (2R)^2} \frac{dA_{HS}}{dR}
\end{align}
While this should give the true mean contact density for any given
system, this is not particularly useful, as we have no way of knowing
the contact density on any given sphere.  However, the mean contact
density comes out as an integral, so we can plot the integrand, and
treat it as a somewhat ill-defined local contact density.

\derivation{

  To compute the mean contact density of a system, we need
  \begin{align}
    \frac{d A_{HS}}{d R} &=
    \frac{\partial A_{HS}}{\partial R} \\
    &+
    \int \left(
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
    \frac{d n_3(\mathbf{r}')}{d R}
    +
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
    \frac{d n_2(\mathbf{r}')}{d R}
    + \cdots
    \right) d\mathbf{r}'
  \end{align}

  \begin{align}
    \frac{dn_3(\mathbf{r})}{dR} &= n_2(\mathbf{r})\\
    \frac{dn_2(\mathbf{r})}{dR} &= \int n(\mathbf{r}+\mathbf{r}')
    \delta'(|\mathbf{r'}| - R)\mathbf{dr}'\\
    &= \int n(\mathbf{r}+\mathbf{r}')
    \delta'(r' - R) d\Omega' r'^2 dr'\\
    &= \int 
    \delta(r' - R) d\Omega' \left(\frac{dn(\mathbf{r}+\mathbf{r}')}{dr'}r'^2 + 2n(\mathbf{r}+\mathbf{r}')r' \right) dr'\\
    &= \frac1{R} \int 
    \delta(|\mathbf{r}'| - R)
    \left(\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') +
    2n(\mathbf{r}+\mathbf{r}') \right) \mathbf{dr}' \\
    &= \frac{2}{R}n_2(\mathbf{r}) + \frac{1}{R} \int 
    \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
    \frac{dn_0(\mathbf{r})}{dR} &= \frac{1}{R} \int 
    \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
    \frac{dn_1(\mathbf{r})}{dR} &=
    \frac1{4\pi R}\frac{dn_2(\mathbf{r})}{dR} - n_0(\mathbf{r}) \\
    &= n_0(\mathbf{r}) + \frac{1}{4\pi R^2} \int 
    \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
    \frac{d\mathbf{n}_{V2}}{dR} &= \nabla n_2
  \end{align}

  The total force due to $\Phi_1$ is actually very easy to work out:
  \begin{align}
    \frac{d A_{HS}^{(1)}}{d R} &=
    \int \mathbf{dr}' \left(\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3) \right)
  \end{align}
  where you can see that $\frac{dn_0(\mathbf{r})}{dR}$ is zero for a
  homogeneous system, and small for reasonably smooth density
  distributions.  The next term from $\Phi_2$ is as follows:
  \begin{align}
    \frac{d A_{HS}^{(2)}}{d R} &=
    \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2} - 2 n_1
      \frac{dn_2}{dR} + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot
    }{1-n_3}
    \right) \\
    &= \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      - 2 n_1 \left(\frac2{R}n_2 \frac{dn_0}{dR}\right)
      + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
    \right) \\
    &= \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      - 4 n_0n_2 - 2 n_1\frac{dn_0}{dR}
      + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
    \right) \\
    &= \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    + \frac{3n_0n_2 + \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      + 2 n_1\frac{dn_0}{dR}
      - 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
    \right)
  \end{align}

  And the last term looks like this.  I should perhaps note that this
  term should be replaced by a tensor version if want to examine
  seriously localized density distributions.
  \begin{align}
    A_{HS}^{(3)} &=
    \int \mathbf{dr}'\left(
    (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
    \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2\left( 1 - n_3 \right)^2
    }
    \right) \\
    \frac{d A_{HS}^{(3)}}{d R} &=
    \int \mathbf{dr}'
    \left[
      3(n_2^2 - \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
      \left(\frac2{R}n_2 + \frac{dn_0}{dR}\right)
      - 6n_2 \mathbf{n}_{V2} \cdot \nabla n_2
      \right]
    \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2\left( 1 - n_3 \right)^2
    }
    \\
    &+
    (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) n_2
    \frac{
      (1 - 2(1-n_3)\ln(1-n_3) - (1-n_3))n_3^2( 1 - n_3)^2
      -
      (n_3 + (1-n_3)^2 \ln(1-n_3))(-2 n_3^2( 1 - n_3) + 2n_3(1-n_3)^2)
    }{
      36\pi (n_3^2\left( 1 - n_3 \right)^2)^2
    }
    \\
    &=
    \int \mathbf{dr}'
    \left[
      (n_1n_2 - \mathbf{n}_{V1} \cdot \mathbf{n}_{V2})
      \left(2n_2 + \frac{1}{4\pi}\frac{dn_0}{dR}\right)
      - \frac{1}{2\pi}n_2 \mathbf{n}_{V2} \cdot \nabla n_2
      \right]
    \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      3 n_3^2\left( 1 - n_3 \right)^2
    }
    \\
    &+
    \frac{n_2(n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})}{36\pi}
    \left( \frac{1}{(1-n_3)^2} + 2 \frac{\ln(1-n_3)}{n_3^2(1-n_3)} \right)
  \end{align}

  \begin{align}
    \frac{d A_{HS}}{d R} &=
    \int \mathbf{dr}' \left(
    -\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3)
    + \cdots
    \right)
  \end{align}
}

@@

  First, let's rewrite the free energy in terms of a smaller set of
  weighted densities:
  \begin{align}
    A_{HS} &= \int \left\{
    -n_0 \ln\left( 1 - n_3\right)
    + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2(1-n_3)^2
    }
    \right\}
    \\
    &= \int \left\{
    -\frac{n_2}{4\pi R^2} \ln\left( 1 - n_3\right)
    + \frac{1}{4\pi R} \frac{n_2^2 - \mathbf{n}_{V2} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2(1-n_3)^2
    }
    \right\}
  \end{align}

>>>>

\affiliation{Department of Physics, Oregon State University, Corvallis, OR 97331}
\author{Jeff Schulte}
\author{Chris Haglund}
\author{Patrick Kreitzburg}
\author{David Roundy}

@@

There has been considerable recent interest in using Statistical
Associating Fluid Theory (SAFT) to construct classical density
functionals to describe associating
fluids\cite{\saftlocaldft,\saftnonlocaldft}.  This approach has been
successful in qualitatively describing the dependence of surface
tension on temperature.  Unfortunately, most of these constructed
functionals\cite{\saftlocaldft} cannot be used to study arbitrary
inhomogeneous density distributions, because they use local density
approximations that are only valid for densities that vary slowly over
molecular length scales.  A key input to those SAFT-based functionals
that can handle sharply varying densities is the correlation function
evaluated at contact. \fixme{I need to figure out how best to phrase
  this... do we want to compute correlation at contact, or contact
  density itself? I prefer the latter, but the former is what is
  commonly included in formulas.}

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  The chain and association contributions to the SAFT
free energy are of particular interest for this paper, since each uses
as input the contact value of the correlation function.
%
Yu and Wu introduced in 2002 a functional for the association term of
the free energy, which included a functional for the contact value of
the correlation function (see
Section~\ref{sec:yuwu})\cite{yu2002fmt-dft-inhomogeneous-associating},
which has subsequently been used in other
functionals\cite{fu2005vapor-liquid-dft, bryk2006density}.
%% Each of these terms has some dependence on the
%% inhomogeneity in the density.  The ideal gas is purely local, and in
%% fact must be the only purely local term in the free energy in order
%% for the functional to satisfy the contact-value theorem.  The
%% hard-sphere contribution contribution to the free energy has been
%% thoroughly studied, and is well approximated by the White Bear version
%% of the Fundamental-Measure Theory (FMT)
%% functional\cite{roth2002whitebear}.
Two functionals for the chain contribution have recently been
introduced\cite{bryk2006density, gross2009density}, one which uses the
contact value of the correlation function of Yu and
Wu\cite{bryk2006density}, while the other introduces a new functional
for the contact value of the correlation
function (see Section~\ref{sec:gross})\cite{gross2009density}.
%% The dispersion term is long-range and thus has significant
%% dependence on the density distribution, but because its relatively
%% weak position dependence, is amenable to mean-field approximations.
Given these different approaches, it seems worthwhile to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.

\fixme{Should we include these equations? Maybe just later?}
\begin{align}
  \frac{A_\textit{chain}}{kT} &= -(m-1) n \left(\ln\left(g_\sigma\right)-1\right)
\end{align}
\begin{align}
  \frac{A_\textit{assoc}}{kT} &= \sum_i n \left(\ln X_i - \frac12 X_i + \frac12\right) \\
  X_i &= \frac{1}{1 + \sum_j n g_\sigma X_j\kappa_{ij} \left(e^{\beta \epsilon_{ij}}-1\right)}
\end{align}

\section{Contact density and correlation function}

\fixme{We should decide on good names for these two densities.  In
  this section I define them formally, and we should keep using the
  same names (and mathematical symbols) for them throughout.}

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
$2R$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
%
\fixme{The following bit is awkward.}
%
In a system with an inhomogeneous density, we seek a \emph{local}
contact value of the correlation function.  There are two reasonable
options for defining such a local function:  a symmetric and an
asymmetric formulation.

The first option treats the two particles that are in contact
symmetrically:
\begin{align}
  g^a_\sigma(\rr) &= \frac{n^a_\textit{contact}(\rr)}{n_a(\rr)}
\end{align}
\begin{align}
  n^a_\textit{contact}(\rr) &= \frac1{4\pi R^2}\int n^{(2)}(\rr - \rr', \rr + \rr')\delta(|\rr'| -
  R)d\rr'
\end{align}
\begin{align}
  n_a(\rr) &= \frac{1}{4\pi R^2}\int n(\rr')\delta(|\rr-\rr'|-R) d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_a$ is the same as the density $n_0$
defined in Fundamental Measure Theory (FMT), and seems ideal for
treating touching spheres.

The second option is to treat one of the two touching spheres
specially:
\begin{align}
  g^b_\sigma(\rr) &= \frac{n^b_\textit{contact}(\rr)}{n_b(\rr)} \\
  n^b_\textit{contact}(\rr) &= \frac1{16\pi R^2}\int n^{(2)}(\rr, \rr + \rr')
  \delta(|\rr'| - 2R)d\rr' \\
  n_b(\rr) &= \frac1{16\pi R^2}\int n(\rr') \delta(|\rr-\rr'|-2R) d\rr'
\end{align}
In this case, we find the average number of contacts for a sphere at a
given position.

In the process of defining these two averaged correlation functions,
we also defined four averaged densities, two \emph{contact
  densities}, $n^a_\textit{contact}(\rr)$ and $n^b_\textit{contact}(\rr)$,
and two \emph{contacting densities}, $n_a(\rr)$ and $n_b(\rr)$.  The
former are the density of spheres that are in contact with each other,
while the latter are the density of spheres available to be in
contact.  In the homogeneous limit, $n_a = n_b = n$ and
$n^a_\textit{contact} = n^b_\textit{contact} = g(\sigma)n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

Yu and Wu introduce an approximation for
$g^a$\cite{yu2002fmt-dft-inhomogeneous-associating}, which will be
discussed in Section~\ref{sec:yuwu}.  Gross introduced an
approximation for $g^b$\cite{gross2009density}, which will be
introduced in Section~\ref{sec:gross}.  We will also introduce our own
approximation to $g^b$ which is based on the assumption of
thermodynamic consistency within FMT, in
Section~\ref{contact-at-sphere}.  We will report on the accuracy of
these approximations by comparing with Monte Carlo simulations of the
inhomogeneous hard-sphere fluid.  We will focus on the contact
densities, rather than the correlation function itself, because these
are simpler, and because they are the quantity that you want to
``average''.  i.e. if you were to average $g$ itself, you'd want to
weight that average based on the density, which would mean averaging
$n^{(2)}$.

@@

%% \section{Direct correlation function}

%% \textcolor{red}{FIXME WRITE THIS SECTION!} The direct correlation
%% function is readily accessible from density-functional theory, as the
%% second functional derivative of the free energy:
%% \textcolor{red}{(Something like...)}
%% \begin{align}
%%   -c^{(2)}(\xx_1, \xx_2) &= \frac{1}{k_BT}\frac{\delta^2 F_{ex}}{\delta n(\xx_1) \delta n(\xx_2)}
%% \end{align}
%% The direct correlation function is related to the ordinary correlation
%% function by the Ornstein-Zernike relation
%% \begin{align}
%%   h(\xx_1, \xx'_2) &= g(\xx_1, \xx_2) - 1 \\
%%   h(\xx_1, \xx_2) &= c(\xx_1, \xx_2)
%%     + \int n(\xx_3)c(\xx_1, \xx_3) h(\xx_2, \xx_3)d\xx''
%% \end{align}
%% From these two equation, we could in principle extract the correlation
%% function evaluated at contact from the DFT free energy functional.
%% The Ornstein-Zernike equation is usually solved by taking a Fourier
%% transform of all the relevant quantities.  In the inhomogeneous case,
%% however, this approach isn't workable, since the Fourier transforms
%% don't speed things up like you'd hope, and we don't have a periodic
%% system in general.

%% \textcolor{red}{It seems worth looking into this, to see what (if
%%   anything) we can do! I don't see a solution just now, but that
%%   doesn't mean there isn't one around somewhere.  In any case, just
%%   looking at the direct correlation function might give us some
%%   hints as to a different and interesting approximation for the
%%   contact density.}

@@

%% \section{Mean Contact density}\label{simple-contact}

%% One reasonably simple question to ask is what the \emph{mean} contact
%% density of an inhomogeneous system is.  This comes down to simply
%% reproducing the derivation for a homogeneous system using the
%% FMT functional.
%% \begin{align}
%%   p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
%%   \ncontact &= \frac{1}{N k_BT 4\pi (2R)^2} \frac{dA_{HS}}{dR}
%% \end{align}
%% While this should give the true mean contact density for any given
%% system, this is not particularly useful, as we have no way of knowing
%% the contact density on any given sphere.  However, the mean contact
%% density comes out as an integral, so we can plot the integrand, and
%% treat it as a somewhat ill-defined local contact density.

%% \derivation{

%%   To compute the mean contact density of a system, we need
%%   \begin{align}
%%     \frac{d A_{HS}}{d R} &=
%%     \frac{\partial A_{HS}}{\partial R} \\
%%     &+
%%     \int \left(
%%     \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}
%%     \frac{d n_3(\mathbf{r}')}{d R}
%%     +
%%     \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}
%%     \frac{d n_2(\mathbf{r}')}{d R}
%%     + \cdots
%%     \right) d\mathbf{r}'
%%   \end{align}

%%   \begin{align}
%%     \frac{dn_3(\mathbf{r})}{dR} &= n_2(\mathbf{r})\\
%%     \frac{dn_2(\mathbf{r})}{dR} &= \int n(\mathbf{r}+\mathbf{r}')
%%     \delta'(|\mathbf{r'}| - R)\mathbf{dr}'\\
%%     &= \int n(\mathbf{r}+\mathbf{r}')
%%     \delta'(r' - R) d\Omega' r'^2 dr'\\
%%     &= \int 
%%     \delta(r' - R) d\Omega' \left(\frac{dn(\mathbf{r}+\mathbf{r}')}{dr'}r'^2 + 2n(\mathbf{r}+\mathbf{r}')r' \right) dr'\\
%%     &= \frac1{R} \int 
%%     \delta(|\mathbf{r}'| - R)
%%     \left(\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') +
%%     2n(\mathbf{r}+\mathbf{r}') \right) \mathbf{dr}' \\
%%     &= \frac{2}{R}n_2(\mathbf{r}) + \frac{1}{R} \int 
%%     \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
%%     \frac{dn_0(\mathbf{r})}{dR} &= \frac{1}{R} \int 
%%     \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
%%     \frac{dn_1(\mathbf{r})}{dR} &=
%%     \frac1{4\pi R}\frac{dn_2(\mathbf{r})}{dR} - n_0(\mathbf{r}) \\
%%     &= n_0(\mathbf{r}) + \frac{1}{4\pi R^2} \int 
%%     \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
%%     \frac{d\mathbf{n}_{V2}}{dR} &= \nabla n_2
%%   \end{align}

%%   The total force due to $\Phi_1$ is actually very easy to work out:
%%   \begin{align}
%%     \frac{d A_{HS}^{(1)}}{d R} &=
%%     \int \mathbf{dr}' \left(\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3) \right)
%%   \end{align}
%%   where you can see that $\frac{dn_0(\mathbf{r})}{dR}$ is zero for a
%%   homogeneous system, and small for reasonably smooth density
%%   distributions.  The next term from $\Phi_2$ is as follows:
%%   \begin{align}
%%     \frac{d A_{HS}^{(2)}}{d R} &=
%%     \int \mathbf{dr}' \left(
%%     \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
%%     n_2
%%     - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2} - 2 n_1
%%       \frac{dn_2}{dR} + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot
%%     }{1-n_3}
%%     \right) \\
%%     &= \int \mathbf{dr}' \left(
%%     \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
%%     n_2
%%     - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
%%       - 2 n_1 \left(\frac2{R}n_2 \frac{dn_0}{dR}\right)
%%       + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
%%     \right) \\
%%     &= \int \mathbf{dr}' \left(
%%     \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
%%     n_2
%%     - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
%%       - 4 n_0n_2 - 2 n_1\frac{dn_0}{dR}
%%       + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
%%     \right) \\
%%     &= \int \mathbf{dr}' \left(
%%     \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
%%     n_2
%%     + \frac{3n_0n_2 + \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
%%       + 2 n_1\frac{dn_0}{dR}
%%       - 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
%%     \right)
%%   \end{align}

%%   And the last term looks like this.  I should perhaps note that this
%%   term should be replaced by a tensor version if want to examine
%%   seriously localized density distributions.
%%   \begin{align}
%%     A_{HS}^{(3)} &=
%%     \int \mathbf{dr}'\left(
%%     (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
%%     \frac{
%%       n_3 + (1-n_3)^2 \ln(1-n_3)
%%     }{
%%       36\pi n_3^2\left( 1 - n_3 \right)^2
%%     }
%%     \right) \\
%%     \frac{d A_{HS}^{(3)}}{d R} &=
%%     \int \mathbf{dr}'
%%     \left[
%%       3(n_2^2 - \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
%%       \left(\frac2{R}n_2 + \frac{dn_0}{dR}\right)
%%       - 6n_2 \mathbf{n}_{V2} \cdot \nabla n_2
%%       \right]
%%     \frac{
%%       n_3 + (1-n_3)^2 \ln(1-n_3)
%%     }{
%%       36\pi n_3^2\left( 1 - n_3 \right)^2
%%     }
%%     \\
%%     &+
%%     (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) n_2
%%     \frac{
%%       (1 - 2(1-n_3)\ln(1-n_3) - (1-n_3))n_3^2( 1 - n_3)^2
%%       -
%%       (n_3 + (1-n_3)^2 \ln(1-n_3))(-2 n_3^2( 1 - n_3) + 2n_3(1-n_3)^2)
%%     }{
%%       36\pi (n_3^2\left( 1 - n_3 \right)^2)^2
%%     }
%%     \\
%%     &=
%%     \int \mathbf{dr}'
%%     \left[
%%       (n_1n_2 - \mathbf{n}_{V1} \cdot \mathbf{n}_{V2})
%%       \left(2n_2 + \frac{1}{4\pi}\frac{dn_0}{dR}\right)
%%       - \frac{1}{2\pi}n_2 \mathbf{n}_{V2} \cdot \nabla n_2
%%       \right]
%%     \frac{
%%       n_3 + (1-n_3)^2 \ln(1-n_3)
%%     }{
%%       3 n_3^2\left( 1 - n_3 \right)^2
%%     }
%%     \\
%%     &+
%%     \frac{n_2(n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})}{36\pi}
%%     \left( \frac{1}{(1-n_3)^2} + 2 \frac{\ln(1-n_3)}{n_3^2(1-n_3)} \right)
%%   \end{align}

%%   \begin{align}
%%     \frac{d A_{HS}}{d R} &=
%%     \int \mathbf{dr}' \left(
%%     -\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3)
%%     + \cdots
%%     \right)
%%   \end{align}
%% }

@@

  First, let's rewrite the free energy in terms of a smaller set of
  weighted densities:
  \begin{align}
    A_{HS} &= \int \left\{
    -n_0 \ln\left( 1 - n_3\right)
    + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2(1-n_3)^2
    }
    \right\}
  \end{align}
